\chapter{Curvature of surfaces}
\section{Example: surfaces of revolution}
Take a curve \(C\) in the plane, and rotate it around a line in that plane (the \emph{axis of revolution}) to create a \emph{surface of revolution}\define{surface!of revolution} \(S\).
Suppose for simplicity that the curve doesn't touch the axis of revolution.
Place the axis of revolution vertically.
Write the points of \(\E[3]\) in cylindrical coordinates \((r \cos \theta, r \sin \theta, z)\).
Our plane is \(\theta=0\), and our axis of revolution is \(r=0\).
Our plane curve \(C\) is arclength parameterized as \(\ot{x}(s)=(r(s),0,z(s))\).
Take adapted frames \(\pr{\ot{x},\ot{e}}\) for the curve in the plane, but swap \(\ot{e}_2\) and \(\ot{e}_3\) so that \(\ot{e}_2\) is normal to the plane and \(\ot{e}_3\) tangent to the plane but normal to the curve:
\begin{align*}
\ot{e}_1(s)&=\frac{d\ot{x}}{ds} = (\cos \varphi(s),0,\sin \varphi(s)),\\
\ot{e}_2(s)&=(0,1,0),\\
\ot{e}_3(s)&=(-\sin \varphi(s),0,\cos \varphi(s)).
\end{align*}

Let \(\rho(\theta)\) be the rotation matrix rotating \(\E[3]\) about the axis of revolution by an angle \(\theta\).
Make an adapted frame at each point of  the surface as 
\begin{align*}
x(s,\theta)&=\rho(\theta)\ot{x}(s)=(r \cos \theta,r\sin\theta,z), \\
e(s,\theta)&=\rho(\theta)\ot{e}(s),
\end{align*}
\begin{problem}{moving.frame:findom12}
Find all \(\omega_i\), \(\gamma_{ij}\).
\end{problem}
\begin{answer}{moving.frame:findom12}
Note that \(\rho\) rotates the point \(\ot{x}\) of our plane around on a circle perpendicular to \(e_1, e_3\) and tangent to \(e_2\), with radius \(r\), so that
\[
\frac{d \rho}{d \theta} \ot{x} = r e_2.
\]
Therefore
\begin{align*}
\omega_1 
&=
e_1 \cdot dx, 
\\
&=
e_1 \cdot d\pr{\rho \ot{x}},
\\
&=
e_1 \cdot \rho \, d\ot{x} + e_1 \cdot \frac{d\rho}{d\theta} \, \ot{x} \, d\theta,
\\
&=
e_1 \cdot \rho \, \ot{e}_1 ds +  e_1 \cdot r e_2 \, d\theta,
\\
&=
e_1 \cdot e_1 ds
\\
&=ds.
\end{align*}
Similarly, \(\omega_3 = 0\) and
\begin{align*}
\omega_2
&=
e_2 \cdot dx, 
\\
&=
e_2 \cdot d\pr{\rho \ot{x}},
\\
&=
e_2 \cdot \rho \, d\ot{x} + e_2 \cdot \frac{d\rho}{d\theta} \, \ot{x} \, d\theta,
\\
&=
e_2 \cdot \rho \, \ot{e}_1 + e_2 \cdot r e_2 \, d\theta,
\\
&=
r \, d\theta.
\end{align*}
Note that 
\begin{align*}
d\ot{e}_1&=\dot\varphi\,\ot{e}_3 ds,\\
d\ot{e}_2&=0,\\
d\ot{e}_3&=-\dot\varphi\,\ot{e}_1 ds,
\end{align*}
and
\[
d\rho=\rho
\begin{pmatrix}
0 & -1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 0
\end{pmatrix}d\theta.
\]
Compute
\begin{align*}
de_1
&=
d(\rho \ot{e}_1),
\\
&=
d\rho \, \ot{e}_1 + \rho \, d\ot{e}_1,
\\
&=
\rho
\begin{pmatrix}
0 & -1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 0
\end{pmatrix}
\ot{e}_1 d\theta
+
\rho \dot{\varphi} \ot{e}_3 ds,
\\
&=
\rho
\begin{pmatrix}
0 & -1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 0
\end{pmatrix}
\begin{pmatrix}
\cos\varphi\\
0\\
\sin\varphi
\end{pmatrix}
d\theta
+
\dot\varphi e_3 ds,
\\
&=
\rho
\begin{pmatrix}
0\\
\cos\varphi\\
0
\end{pmatrix}
d\theta
+
\dot\varphi e_3 ds,
\\
&=
\cos\varphi e_2d\theta
+
\dot\varphi e_3 ds.
\end{align*}
and similarly
\begin{align*}
de_2
&=
d(\rho \ot{e}_2),
\\
&=
d\rho \, \ot{e}_2,
\\
&=
\rho
\begin{pmatrix}
0 & -1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 0
\end{pmatrix}
\ot{e}_2 d\theta
\\
&=
-
\rho
\begin{pmatrix}
1 \\
0 \\
0
\end{pmatrix}
d\theta
\\
&=
-\begin{pmatrix}
\cos\theta\\
\sin\theta\\
0
\end{pmatrix}
d\theta.
\end{align*}
Finally,
\begin{align*}
de_3
&=
d(\rho \ot{e}_3),
\\
&=
d\rho \, \ot{e}_3 + \rho \, d\ot{e}_3,
\\
&=
\rho
\begin{pmatrix}
0 & -1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 0
\end{pmatrix}
\ot{e}_3 d\theta
-
\rho \dot\varphi \ot{e}_1 ds,
\\
&=
\rho
\begin{pmatrix}
0 & -1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 0
\end{pmatrix}
\begin{pmatrix}
-\sin\varphi\\
0\\
\cos\varphi
\end{pmatrix}
d\theta
-
\dot\varphi e_1 ds,
\\
&=
\rho
\begin{pmatrix}
0\\
-\sin\varphi\\
0
\end{pmatrix}
d\theta
-
\dot\varphi e_1 ds,
\\
&=
-\sin\varphi \, e_2d\theta
-\dot\varphi e_1 ds.
\end{align*}
Therefore 
\begin{align*}
\gamma_{12}&=-\gamma_{21},\\
&=-e_2\cdot de_1,\\
&=-e_2\cdot(\cos\varphi \, e_2 d\theta+\dot\varphi e_3 ds),\\
&=-\cos\varphi \, d\theta.
\end{align*}
\begin{align*}
\gamma_{13}&=e_1 \cdot de_3,\\
&=-\dot\varphi \, ds,\\
\end{align*}
\begin{align*}
\gamma_{23}&=e_2\cdot de_3,\\
&=-\sin\varphi \, d\theta.
\end{align*}
\end{answer}
The shape operator:
\[
\begin{pmatrix}
\gamma_{13}\\
\gamma_{23}
\end{pmatrix}
=
\begin{pmatrix}
-\dot\varphi & 0 \\
0 & -\frac{\sin\varphi}{r}
\end{pmatrix}
\begin{pmatrix}
\omega_1\\
\omega_2
\end{pmatrix}.
\]
So the Gauss curvature of a surface of revolution is
\[
K=\frac{\dot\varphi \sin \varphi}{r}=-\frac{\ddot{r}}{r},
\]
while the mean curvature is
\[
H=-\frac{1}{2}\pr{\dot\varphi+ \frac{\sin \phi}{r}}.
\]
\begin{example}
Take a function \(H(s)\) and try to construct a curve \(\ot{x}(s)\) in the plane, so that the associated surface of revolution has mean curvature \(H(s)\).
To do this, we have to solve the equations
\begin{align*}
\dot{r}(s) &= \cos \varphi(s), \\
\dot\varphi(s) &= -2H(s)-\frac{\sin \varphi(s)}{r(s)}
\end{align*}
with \(r > 0\).
Each initial value \(r=r_0>0\) and \(\varphi=\varphi_0\) at some \(s=s_0\) determines a unique local solution, by the existence and uniqueness theorem for ordinary differential equations.
Solve for \(z\) by \(z(s) = \int \sin \varphi(s) \, ds\), again unique up to adding a constant.
\end{example}
\begin{problem}{moving.frame:integral}
If we want \(H\) to be constant, say equal to a constant \(H_0\), can you integrate this coupled system of ordinary differential equations?
Hint: is \(r \sqrt{1-\dot{r}^2}+H_0 r^2\) constant along solutions?
\end{problem}
\begin{answer}{moving.frame:integral}
If \(\sin\varphi=0\) everywhere, the curve is a horizontal line and the surface of revolution an annulus.
Suppose that \(\sin\varphi\) is not everywhere zero; pick an interval on which \(\sin\varphi\ne 0\), and let \(\varepsilon\) be the sign of \(\sin\varphi\).
Differentiate to find that the function 
\[
\beta\defeq\varepsilon r \sqrt{1-\dot{r}^2}+H_0 r^2
\]
is constant along solutions, say equal to \(\beta_0\).
Solve for \(\dot{r}\):
\[
\dot{r}^2 = 1-\frac{\beta_0-H_0r^2}{r^2}.
\]
Substitute \(u=r^2\):
\[
\dot{u}=2\sqrt{H_0u^2+u-\beta_0}.
\]
Integrate:
\[
\int\frac{du}{2\sqrt{H_0u^2+u-\beta_0}}=s.
\]
This integral can be solved in elementary functions giving \(s=s(r)\); for example, if \(H_0>0\),
\begin{align*}
s&=
\frac{
	(4\beta_0H_0+1)
	\log
	\pr{
		2\sqrt{H_0}
		{
			\sqrt{
				H_0u^2+u-\beta_0
				}
			+2H_0u+1
		}
	}
	}{16H_0^{3/2}},
\\
&=
\frac{
	(4\beta_0H_0+1)
	\log
	\pr{
		2\sqrt{H_0}
		{
			\sqrt{
				H_0r^4+r^2-\beta_0
				}
			+2H_0r^2+1
		}
	}
	}{16H_0^{3/2}},
\end{align*}
The surfaces of revolution of constant positive mean curvature are obtained by solving implicitly for \(r=r(s)\).
\end{answer}
\begin{example}
Similarly, we can prescribe Gauss curvature \(K(s)\) arbitrarily by solving the equation \(\ddot{r}+Kr = 0\).
For example, the surfaces of revolution with zero Gauss curvature are the circular cones given by rotating a line segment.
More generally, for constant Gauss curvature \(K=K_0\), the solutions are (with \(k_0\defeq\sqrt{\left|K_0\right|}\))
\[
r(s) = \begin{cases}
a\cos k_0s+ b\sin k_0s, & \text{ if \(K_0 > 0\)}, \\
a+bs, & \text{ if \(K_0=0\)}, \\
a\cosh k_0 s
+
b\sinh k_0 s, & \text{ if \(K_0 < 0\)}.
\end{cases}
\] 

For \(K_0>0\),  translate \(s\) to arrange \(b=0\), so write
\[
r(s)=r_0 \cos k_0 s,
\]
so that
\[
\dot{z}^2=1-\dot{r}^2=1-r_0^2K_0 \sin^2 k_0 s,
\]
giving
\[
z= \int \sqrt{1-r_0^2K_0 \sin^2 k_0 s} \, ds,
\]
an elliptic integral, usually denoted \(E(k_0s \, | \, r_0^2K_0)/k_0\).
Similar remarks apply to \(K_0<0\).
\end{example}
\begin{problem}{moving.frames:Weingarten}
Take a curve \(C_0\) in the plane \(\R[2]\).
Prove that there is a curve \(C_1\) in the plane \(\E[2]\) whose associated surface of revolution satisfies \((H,K) \in C_0\) at every point.
\end{problem}

\section{Orientations}
We could have used only positively oriented frames \(e_1,e_2,e_3\) of \(\E[3]\).
We could then look at an oriented surface, and define its \emph{adapted frames} to have \(e_1,e_2\) a positively oriented basis of the tangent plane, and \(e_1,e_2,e_3\) a positively oriented basis of \(\E[3]\).
The expression 
\[
g=
\begin{pmatrix}
h & 0 \\
0 & n
\end{pmatrix}
\]
simplifies to having \(h\) a rotation matrix, and \(n=1\), so
\[
g=
\begin{pmatrix}
\cos \theta & -\sin \theta & 0 \\
\sin \theta & \cos \theta & 0 \\
0 & 0 & 1
\end{pmatrix},
\]
which we just write as \(e^{i\theta}\).
\begin{problem}{moving.frame:area.defined}
Over an oriented surface, prove that the \(2\)-form \(dA\defeq\omega_1\wedge\omega_2\) is the pullback of a unique \(2\)-form \(dA\) on the surface, the \emph{area form}\define{area form}.
\end{problem}
\begin{problem}{moving.frame:mean.curv.defined}
Over an oriented surface, prove that the mean curvature is the pullback of a unique function on the surface.
\end{problem}
\begin{problem}{moving.frame:Hopfy}
Prove that, on any surface, oriented or not, the shape operator admits a unique expression as 
\[
\frac{q}{2} \omega^2 + \frac{H}{2} (\omega \bar\omega + \bar\omega\omega) + \frac{\bar{q}}{2} \bar\omega^2.
\]
for functions \(q \colon \frameBundle{S} \to \C\),  \(H \colon \frameBundle{S} \to \R\).
How are the \(a_{ij}\) related to \(q,H\)?
\end{problem}
\begin{answer}{moving.frame:Hopfy}
Taking real and imaginary parts \(q=q_1+iq_2\):
\[
\begin{pmatrix}
a_{11} & a_{12} \\
a_{12} & a_{22}
\end{pmatrix}
=
\begin{pmatrix}
H+q_1 & -q_2 \\
-q_2 & H-q_1
\end{pmatrix}.
\]
So \(H=(a_{11}+a_{22})/2\) is the mean curvature, and
\[
q=\frac{a_{11}-a_{22}-2i\, a_{12}}{2}.
\]
Note that \(|q|^2=q\bar{q}=H^2-K\), so \(q=0\) precisely at umbilic points.
\end{answer}
The expression \(q\omega^2\) is the \emph{Hopf differential}.\define{Hopf differential}
\begin{problem}{moving.frame:Hopf.2}
If the surface is oriented, prove that the Hopf differential descends to the surface, as a complex valued quadratic form on tangent vectors.
\end{problem}
The \emph{Gauss map}\define{Gauss map} of a surface is \(e_3\), mapping the frame bundle to the unit sphere.
\begin{problem}{moving.frame:Gauss.map}
Prove that the Gauss map of an oriented surface is the composition of the map \((x,e)\in\frameBundle{S}\mapsto x\) with a unique map \(x \in S\mapsto e_3(x)\), also called the \emph{Gauss map}.
\end{problem}


\optionalSection{Recovering a surface from its inner products and shape operator}
\begin{theorem}[Bonnet]\define{theorem!Bonnet}\define{Bonnet!theorem}
If two connected oriented surfaces in \(3\)-dimensional Euclidean space are identified by  a diffeomorphism preserving orientation and the inner product and shape operator on every tangent space, then the diffeomorphism is induced by an orientation preserving rigid motion of \(3\)-dimensional Euclidean space.
\end{theorem}
\begin{proof}
Use the orientations of the surfaces to define ``identifying'' the shape operators: we have a well defined normal vector on each surface, so we treat the shape operator as a symmetric bilinear form on tangent spaces.
Such a diffeomorphism \(S \to \ot{S}\), clearly identifies positively oriented orthonormal frames \(e_1,e_2\) of tangent spaces of \(S\) with those of \(\ot{S}\).
Taking the usual orientation of \(\E[3]\), these extend uniquely into positive oriented orthonormal frames \(e_1,e_2,e_3=e_1\times e_2\), a diffeomorphism of the adapted frames, clearly identifying the soldering \(1\)-forms \(\omega_1,\omega_2\) and \(\omega_3=0\).
Differentiate to find that the Levi--Civita connection \(1\)-form \(\gamma_{12}\) matches.
Having the same shape operator ensures precisely that \(\gamma_{31},\gamma_{32}\) match.
It then follows by antisymmetry of \(\gamma_{ij}\) that all connection \(1\)-forms agree.
Let 
\[
h=h(x,e)=
\begin{pmatrix}
1 & 0\\
x & e
\end{pmatrix}
\]
as above, and \(\ot{h}\) in the same way for the corresponding surface:
\[
h^{-1}dh=\ot{h}^{-1}d\ot{h}=
\begin{pmatrix}
0 & 0 \\
\omega & \alpha
\end{pmatrix},
\]
under the diffeomorphism.
So \(h\) and \(\ot{h}\) satisfy the same first order linear ordinary differential equation 
\[
dh = 
\begin{pmatrix}
0 & 0 \\
\omega & \alpha
\end{pmatrix}
h
\]
along any curve in the frame bundle \(\frameBundle{S}\).
Apply a rigid motion to one of the two surfaces, so that their frame bundles contain at least one frame in common.
By existence and uniqueness for linear ordinary differential equations, \(h=\ot{h}\) along any curve in \(\frameBundle{S}\) starting at that frame.
The adapted frames which are positively oriented for \(\E[3]\) and with \(e_1,e_2\) positively oriented on the surface form a connected manifold.
Hence \(h=\ot{h}\).
\end{proof}

\optionalSection{Parallel surfaces}\SubIndex{parallel surfaces}%
Start with a surface \(S\) with a chosen normal direction, i.e. a unit normal vector \(e_3\) at each point, smoothly varying.
Let \(\ot{S}\) be the translation of \(S\) in that direction by a distance \(t\).
So then \(\ot{x}=x+te_3\), \(d\ot{x}=dx+t \, de_3\) and \(e_3\cdot d\ot{x}=e_3 \cdot dx+te_3\cdot de_3=0\), so \(e_3\) is still normal to \(\ot{S}\) at \(\ot{x}\).
Hence \(T_{\ot{x}} \ot{S} = T_x S\) is constant in \(t\), and the Gauss map is unchanged.
Identify \(\frameBundle{\ot{S}}\cong\frameBundle{S}\) by 
\[
\Phi \colon (x,e)\mapsto (\ot{x},\ot{e})=(x+te_3,e).
\]
The soldering \(1\)-forms are
\begin{align*}
\otomega_i&=e_i\cdot d\ot{x},
\\
&=e_i\cdot dx + t e_i \cdot de_3,
\\
&=
\omega_i+t\gamma_{i3},
\\
&=
\omega_i+ta_{ij}\omega_j,
\end{align*}
which we can write as \(\otomega=(I+ta)\omega\).
The Levi-Civita connection \(1\)-form is
\[
\qc_{12}=e_1\cdot de_2=\gamma_{12}.
\]
The shape operator is
\begin{align*}
\qc_{i3}&=e_i\cdot de_3,\\
&=\gamma_{i3},
\\
&=
a_{ij}\omega_j
\end{align*}
If we write this as \(\qc_{*3}=a\omega\), then 
\[
\qc_{*3}=a\omega=a(I+ta)^{-1}\qf{},
\]
so
\[
\ot{a}=a(I+ta)^{-1}.
\]
In any frame in which \(a\) is diagonalized, so is \(\ot{a}\):
\[
a=
\begin{pmatrix}
k_1&0\\
0&k_2
\end{pmatrix}
\]
in principal curvatures, so 
\[
\ot{a}=
\begin{pmatrix}
\ot{k}_1&0\\
0&\ot{k}_2
\end{pmatrix}
=
\begin{pmatrix}
\frac{k_1}{1+tk_1}&0\\
0&\frac{k_2}{1+tk_2}
\end{pmatrix}.
\]
Hence Gauss curvature
\[
\ot{K}=\frac{\det a}{\det(I+ta)}=\frac{K}{1+2Ht+Kt^2}
\]
and mean curvature
\[
\ot{H}=\frac{H+Kt}{1+2Ht+Kt^2}.
\]
In particular, the parallel surface will not be smooth enough to compute such quantities at any point \(\ot{x}=x+te_3\) where \(1+tk_1\) or \(1+tk_2\) vanish.
But if these don't vanish, then the differential of the map is
\[
d\ot{x}=dx+t \, de_3= \omega_i \, e_i + ta_{ij} \omega_je_i
\]
an immersion.
Surfaces parallel to a Weingarten surface are also Weingarten, but for a different relation between mean and Gauss curvature.

\optionalSection{Constant mean curvature surfaces}%
On an oriented surface, a \emph{complex differential}\define{complex differential} is an expression of the form \(f\omega^k\) on the frame bundle, but which is in fact defined on the surface.
In other words, it doesn't transform as we right translate by \(e^{i\theta}\).
Since \(\omega\) transforms by \(e^{-i\theta}\), we need \(f\) to transform by \(e^{ik\theta}\):
\[
r_{e^{i\theta}}^*f=e^{ik\theta}f.
\]
\begin{problem}{moving.frame:diff.Hopf}
Differentiating this, show that
\[
df+ikf\alpha=Df \omega + \bar{D}f \bar\omega,
\]  
for unique complex functions \(Df,\bar{D}f\).
\end{problem}
\begin{answer}{moving.frame:diff.Hopf}
In our complex notation, we are identifying the matrix
\[
\begin{pmatrix}
0 & -1 & 0 \\
1 &  0 & 0 \\
0 & 0 & 0
\end{pmatrix}
\]
with the complex number \(i\).
The fibers of \(\frameBundle{S} \to S\) have the form
\[
(x(t),e(t))=(x_0,e_0e^{it}).
\]
So if we write out \(\gamma=\transpose{e}de\) in our complex notation,
on each fiber,
\[
\gamma=i \, dt,
\]
so \(\alpha=\gamma_{12}=-dt\).
We are given that
\[
f(x(t),e(t))=f(x_0,e_0e^{it})=e^{ikt}f(x_0,e_0),
\]
so on the fiber
\[
\frac{df}{dt}=ikf,
\]
i.e.
\[
df+ikf\alpha=0.
\]
So in complex notation, \(df+ikf\alpha\) vanishes on vertical vectors for \(\frameBundle{S} \to S\), i.e. is a multiple of \(\omega,\bar\omega\).
\end{answer}
A \emph{holomorphic differential}\define{holomorphic differential} is one for which \(0=\bar{D}f\), i.e. \(df-ikf\alpha\) is complex linear.

A \emph{holomorphic chart} on an oriented surface in \(\E[3]\) is a chart on an open subset of the surface with differential \(dz\) a holomorphic differential.
Any two such charts locally agree up to complex analytic transition maps.
The Korn--Lichtenstein theorem\SubIndex{theorem!Korn--Lichtenstein}\SubIndex{Korn--Lichtenstein theorem} says that any oriented surface is covered by domains of holomorphic charts.
Hence any oriented surface in \(\E[3]\) is a Riemann surface.
\begin{problem}{moving.frame:Hopf.1}
Compute that the Hopf differential satisfies
\[
dq+2iq\alpha = Dq \omega +  \bar{D}q \bar\omega,
\]
where
\(
\bar{D}q=
2(H_1-H_2i)
\)
where \(H\) is the mean curvature, and \(dH=H_1\omega_1+H_2\omega_2\).
Thereby, prove:
\end{problem}
\begin{answer}{moving.frame:Hopf.1}
\begin{align*}
Dq&=a_{111}-3a_{122}+i(a_{222}-3a_{112}),\\
\bar{D}q&=2(H_1-iH_2).
\end{align*}
\end{answer}
\begin{theorem}
An oriented surface has constant mean curvature just when its Hopf differential is holomorphic.
\end{theorem}
The Riemann--Roch theorem \cite{Griffiths:1989} p. 99 shows that a compact Riemann surface of genus \(g\) has a space of quadratic differentials of complex dimension: 
\[
\begin{cases}
0,&g=0,\\
1,&g=1,\\ 
3g-3,&g\ge 2.
\end{cases}
\]
It is well known that there is precisely one genus zero compact Riemann surface, up to holomorphic diffeomorphism: the Riemann sphere \cite{Griffiths:1989} p. 125.
Applying Bonnet's theorem:
\begin{corollary}
For any constant \(H_0\), the sphere has at most one immersion in \(\E[3]\) with constant mean curvature equal to \(H_0\), up to rigid motion: as a round sphere of radius \(1/H_0\).
\end{corollary}

\section{Away from umbilics}
Recall that principal curvatures of a surface are the eigenvalues of its shape operator.
Suppose that they are distinct \(k_1\ne k_2\), i.e. there are no umbilic points.
A \emph{Darboux frame}\define{Darboux frame}\define{frame!Darboux} is one in which \(e_1,e_2\) are eigenvectors of the shape operator with eigenvalues \(k_1<k_2\).
The set of Darboux frames of the surface is a surface, as each Darboux frame is uniquely determined up to signs.
On that surface \(\gamma_{i3}=k_i\omega_i\), for \(i=1,2\).
\prob{moving.frame:not.umb}%
{Differentiate the equations of a Darboux frame to see that
\[
d
\begin{pmatrix}
k_1\\
k_2
\end{pmatrix}
=
\begin{pmatrix}
k_{11}&k_{12}\\
k_{21}&k_{22}
\end{pmatrix}
\begin{pmatrix}
\omega_1\\
\omega_2
\end{pmatrix},
\]
for some functions \(k_{ij}\), perhaps \emph{not} symmetric in lower indices, while
\[
\gamma_{12}=\frac{k_{12}\omega_1+k_{21}\omega_2}{k_2-k_1}.
\]}
\begin{answer}{moving.frame:not.umb}%
\begin{align*}
0
&=
\begin{pmatrix}
d(\gamma_{13}-k_1\omega_1)\\
d(\gamma_{23}-k_2\omega_2)
\end{pmatrix}
\\
&=
\begin{pmatrix}
-\gamma_{12}\wedge\gamma_{23}-dk_1\wedge\omega_1+k_1\gamma_{12}\wedge\omega_2\\
-\gamma_{21}\wedge\gamma_{13}-dk_2\wedge\omega_2+k_2\gamma_{21}\wedge\omega_1
\end{pmatrix}
\\
&=
\begin{pmatrix}
-\gamma_{12}\wedge k_2\omega_2-dk_1\wedge\omega_1+k_1\gamma_{12}\wedge\omega_2\\
\gamma_{12}\wedge k_1\omega_1-dk_2\wedge\omega_2-k_2\gamma_{12}\wedge\omega_1
\end{pmatrix}
\\
&=
-%
\begin{pmatrix}
dk_1\wedge\omega_1+(k_2-k_1)\gamma_{12}\wedge\omega_2\\
(k_2-k_1)\gamma_{12}\wedge\omega_1+dk_2\wedge\omega_2
\end{pmatrix}
\end{align*}
\end{answer}
Expand \(0=d^2k_i\) to find that, if we let
\begin{align*}
Dk_{i1}&=dk_{i1}+k_{i2}\gamma_{12},\\
Dk_{i2}&=dk_{i2}-k_{i1}\gamma_{12},
\end{align*}
then
\(
Dk_{ij}=k_{ijk}\omega_k
\)
with \(k_{ijk}=k_{ikj}\).
Recall that Gauss curvature is \(K=k_1k_2\).
Plug into the equation
\[
d\gamma_{12}=K\omega_1\wedge\omega_2.
\]
to get
\[
d\left(
\frac{k_{12}\omega_1+k_{21}\omega_2}{k_2-k_1}
\right)
=
k_1k_2\omega_1\wedge\omega_2.
\]
Expanding this out gives
\[
0=
\frac{k_{122}-k_{211}}{k_2-k_1}
+\frac{2k^2_{12}+2k^2_{21}}{(k_2-k_1)^2}
+k_1k_2.
\]

\begin{lemma}\label{lemma:max.min.k}
If a surface bears a point at which the larger principal curvature is maximal and the smaller minimal, then either the point is umbilic or the Gauss curvature is not positive at that point.
\end{lemma}
\begin{proof}
We can assume that \(k_2>k_1\) near that point, and then adapt frames as above.
At a critical point of \(k_2\) and \(k_1\), \(0=dk_1=dk_2\), so \(k_{ij}=0\) and so \(\gamma_{12}=0\), so \(Dk_{ij}=dk_{ij}=k_{ijk}\omega_k\).
Since \(k_2\) is maximal, along the flow of \(e_1\), \(k_{211}\le 0\).
Since \(k_1\) is minimal, along the flow of \(e_2\), \(k_{122}\ge 0\).
Plug in above to get \(K=k_1k_2\le 0\).
\end{proof}

\begin{theorem}[Liebmann]\SubIndex{Liebmann, H.}
Suppose that \(S\) is a connected compact surface in \(\E[3]\) of (i) constant Gauss curvature or (ii) constant mean curvature with Gauss curvature not having any critical zeroes.
Then \(S\) is a round sphere.
\end{theorem}
\begin{proof}
By problem~\vref{problem:moving.frame:compact}, there is a point of positive Gauss and mean curvature, so the constant is a positive constant.
At any point at which the principal curvatures are as far apart from one another as possible, one is maximal and the other minimal, as their product is constant.
Apply lemma~\vref{lemma:max.min.k} to find a point of nonpositive Gauss curvature where the Gauss curvature is critical.
\end{proof}
\prob{moving.frame:distinct.lambda}{Suppose that \(S\) is a connected surface in \(\E[3]\) and that one principal curvature is a nonzero constant, and always distinct from the other.
Prove that \(S\) is a union of circular arcs of constant radius, with centres on a curve.}
\begin{answer}{moving.frame:distinct.lambda}
Arrange \(e_1,e_2\) to diagonalize the shape operator, so 
\begin{align*}
\gamma_{13}&=k_1\omega_1,\\
\gamma_{23}&=k_2\omega_2,
\end{align*}
with \(k_1\) constant.
Differentiate to find that \(\gamma_{12}\) is a multiple of \(\omega_2\).
So then along the flow of \(e_1\), \(\gamma_{12}=0\) and \(\gamma_{23}=0\) i.e. \(e_2\) is constant, so \(e_1,e_3\) rotate in the plane, with \(\gamma_{13}=k_1\), so at a constant rate, i.e. on a circle.
\end{answer}

We define completeness in appendix~\ref{chapter:geodesics}; for the moment, we only need know that on a complete surface,\SubIndex{complete!surface}\SubIndex{surface!complete} every bounded vector field is complete.\SubIndex{complete!vector field}\SubIndex{vector field!complete} 
\begin{theorem}[Hilbert]\define{theorem!Hilbert}\define{Hilbert!surface theorem}
No complete surface in \(\E[3]\) has constant negative Gauss curvature.
\end{theorem}
\begin{proof}
Take a surface of constant negative Gauss curvature, say equal to \(-1\).
We can assume it is connected, and replace it by its universal covering space if needed, so assume it is simply connected.
Let \(u\) be half the angle between the asymptotic curves, so \(0<u<\pi/2\).
Check that the principal curvatures are \(-\cot u,\tan u\).
A surface of negative Gauss curvature has no umbilics.
Differentiate the equations 
\begin{align*}
\gamma_{13}&=-\cot u \, \omega_1,\\
\gamma_{23}&=\tan u \, \omega_2.
\end{align*}
Check that \(\omega_1/\sin u\) and \(\omega_2/\cos u\) are closed \(1\)-forms; noting that \(0 < u < \pi/2\), so the denominators don't vanish.
Because the surface is connected and simply connected, there are functions \(s_1,s_2\) so that 
\begin{align*}
ds_1&=\omega_1/\sin u,\\
ds_2&=\omega_2/\cos u,
\end{align*}
unique up to constant and with \(ds_1,ds_2\) linearly independent.

The surface is complete, so \(e_1,e_2\) have complete flows.
The indefinite integral \(\int \omega_1\) along curves \(\omega_2=0\) is an unbounded function, defined up to constant.
But \(s_1\) grows at least as fast, so is also unbounded along curves on which \(ds_2=0\), and vice versa.
So the map \((s_1,s_2)\) from our surface to the plane is onto, and a local diffeomorphism.

The vector fields \(\p{s_1}=\sin u \, e_1\), \(\p{s_2}=\cos u \, e_2\) are complete and commuting.
Flows of these for times \(s_1,s_2\) invert our previous map, hence a diffeomorphism.

Let \(x\defeq(s_1-s_2)/2\), \(y\defeq(s_1+s_2)/2\).
The level sets of \(x\) and \(y\) are precisely the asymptotic curves.
Check that \(\gamma_{12}=-u_xdx+u_ydy\), and that the relation \(d\gamma_{12}=K\omega_1\wedge\omega_2\) becomes the \emph{sine--Gordon equation}\SubIndex{sine--Gordon equation} for the angle \(\theta=2u\):
\[
\theta_{xy}=\sin\theta.
\]
Note also that \(0<\theta<\pi\), so \(\sin\theta>0\).
The area form is \(\omega_1\wedge\omega_2=2\sin \theta dx \wedge dy\).

Apply Stokes's theorem to a rectangle \(R\) in the \(xy\)-plane
\[
\int_{\partial R}\gamma_{12}=\int_R d\gamma_{12},
\]
to find
\[
\theta(x,y)-\theta(0,y)=\theta(x,0)-\theta(0,0)+\int \sin \theta \, dx \, dy.
\]
The difference in values of \(\theta\) along the top of any rectangle is at least as large as the difference along the bottom.

Since \(0<\theta<\pi\), the area of the rectangle, in the geometry of our surface, is never more than \(4\pi\).
Our coordinates are global, so any measureable subset of the surface has area no more than \(4\pi\).
Each set of infinite Euclidean area thus contains sets of large measure where \(\sin \theta\) gets small, so \(\theta\) gets close to \(0\) or \(\pi\).

Suppose that \(\theta\) is not constant on some horizontal line.
Change the sign of \(x\) and \(y\) if needed, to arrange that \(\theta\) increases along a horizontal line segment.
Split that line segment into three line segments.
On the leftmost, suppose that \(\theta\) increases by some amount \(\varepsilon_L\), and on the rightmost by some amount \(\varepsilon_R\).
\[
\begin{tikzpicture}[scale=.7]
\fill[gray!40,opacity=.25](-1,0) rectangle (1,1);
\draw[thick,gray] (-1,1) -- (-1,0) -- (1,0) -- (1,1);
\draw[thick,gray] (-.333,1) -- (-.333,0) -- (.333,0) -- (.333,1);
\node (b) at (-.333,-.1){};
\node (c) at (.333,-.1){};
\draw[-stealth,thick,gray] (-1,-.1) -- (-.333,-.1);
\node[below] at (-.666,-.1) {\(\varepsilon_L\)};
\draw[-stealth,thick,gray] (.333,-.1) -- (1,-.1);
\node[below] at (.666,-.1) {\(\varepsilon_R\)};
\end{tikzpicture}
\]
Draw the half strips above the line segments.
So \(\theta\) increases by at least \(\varepsilon_L\) across each parallel line segment across the left half strip.
But \(\theta>0\) everywhere, so in the middle strip, \(\theta>\varepsilon_L\).
Similarly, in the middle strip, \(\theta<\pi-\varepsilon_R\).
So \(\sin \theta\) is bounded away from zero in the middle strip, a contradiction.

Hence \(\theta\) is constant on every horizontal line, i.e. \(0=\theta_x\), so \(0=\theta_{xy}=\sin \theta\), a contradiction.
\end{proof}

